---
title: "量化因子分析报告"
author: "吴松键"
date: "`2026-04-01`"
output: html_document
---

# 1. 因子构建(Factor Construction)

计算基于"低收益日资金流向"的因子(Calculate the factor based on "low-return day fund flow")

```{r, warning = FALSE, message = FALSE,include=FALSE}
library(tidyverse)
library(lubridate)
library(zoo)

# 读取数据 Read data
df <- read.csv("dataset.csv")
df$date <- as.Date(df$date)
df <- df %>% arrange(order_book_id, date)
```

```{r}
# 计算基础指标 a
# Calculate a
df <- df %>%
  mutate(a = (buy_value - sell_value) / (buy_value + sell_value))

# 计算 factor 函数
# Calculate factor function
calc_factor <- function(returns, a_vals) {
  n <- length(returns)
  factor_vec <- rep(NA, n)
  for (i in 20:n) {
    window_ret <- returns[(i-19):i]
    window_a <- a_vals[(i-19):i]
    # 选取收益率最低的 4 天 (20 * 0.2 = 4)
    # Select the 4 days with the lowest return
    idx <- order(window_ret)[1:4]
    factor_vec[i] <- mean(window_a[idx])
  }
  return(factor_vec)
}

# 按股票分组计算
# Calculate by stock group
df <- df %>%
  group_by(order_book_id) %>%
  mutate(factor = calc_factor(pct_rate, a)) %>%
  ungroup() %>%
  drop_na(factor)
```

# 2. IC 分析(IC Analysis)

计算因子与下一期收益率的相关性(Calculate the correlation between the factor and the next period's return)

```{r}
df_analysis <- df %>%
  group_by(order_book_id) %>%
  mutate(next_ret = lead(pct_rate)) %>%
  ungroup() %>%
  drop_na(next_ret)

# 计算每日 IC Calculate daily IC
ic_series <- df_analysis %>%
  group_by(date) %>%
  summarise(ic = cor(factor, next_ret, use = "complete.obs"))

cat("平均 IC 值, Mean IC value:", mean(ic_series$ic, na.rm = TRUE), "\n")
cat("IC IR:", mean(ic_series$ic, na.rm = TRUE) / sd(ic_series$ic, na.rm = TRUE), "\n")
```

# 3. 分组收益图

```{r}
# 分组 Grouping
df_analysis <- df_analysis %>%
  group_by(date) %>%
  mutate(group = ntile(factor, 5)) %>%
  ungroup()

# 计算累积收益 cumulative returns
cum_ret <- df_analysis %>%
  group_by(date, group) %>%
  summarise(daily_ret = mean(next_ret, na.rm = TRUE)) %>%
  group_by(group) %>%
  mutate(cum = cumprod(1 + daily_ret))

ggplot(cum_ret, aes(x = date, y = cum, color = as.factor(group))) +
  geom_line() + theme_minimal() +
  labs(title = "因子分组累计收益图 Cumulative Return by Factor Group", x = "日期Date", y = "累计收益Cumulative Return")
```

```{r}
ggplot(ic_series, aes(x = date, y = ic)) +
  geom_smooth(method = "loess", color = "blue", se = FALSE, alpha = 0.5) +
  geom_line(color = "black", alpha = 0.4) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) +
  theme_minimal() +
  labs(title = "IC 时间序列走势图Time Series Trend",
       subtitle = "蓝色线条为趋势平滑线，红色虚线为零轴 Blue line is the smoothed trend, red dashed line is the zero axis",
       x = "日期 Date", 
       y = "IC 值 IC Value") +
  theme(plot.title = element_text(hjust = 0.5))
```

本次分析发现因子 IC 均值较低（约 0.007），且时间序列图有随机震荡特征。
This analysis found that the mean IC of the factor is low (approximately 0.007), and the time series chart shows random oscillation characteristics.
